---
title: "Lab 06 - Ugly charts and Simpson's paradox"
author: "Cynthia Jiao"
date: "2/17/2025"
output: github_document
---

### Load packages and data

```{r load-packages, message = FALSE}
library(tidyverse) 
library(dplyr)
library(dsbox)
library(mosaicData) 
library(readxl)
```

```{r connect-github}

library(usethis)
use_git_config(
  user.name = "CynthiaJiao",
  user.email = "cynthia.rj.2001@gmail.com"
)
```

### Staff Data

```{r staff-long-data-graphs}

staff <- read_csv("data/instructional-staff.csv")
print(staff)

## converting to long data

staff_long <- staff %>%
  pivot_longer(cols = -faculty_type, names_to = "year") %>%
  mutate(value = as.numeric(value))

staff_long

## line graph for long data!

staff_long %>%
  ggplot(aes(x = year, y = value, color = faculty_type)) +
  geom_line()

## group the observations in faculty type

staff_long %>%
  ggplot(aes(
    x = year,
    y = value,
    group = faculty_type,
    color = faculty_type
  )) +
  geom_line() +
  ggtitle("Trends in Instructional Staff Employment Status, 1975-2011") +
  labs(x = "Year", y = "Percent of Total Instructional Staff", color = "Faculty Types") +
  theme_light()


```

```{r part-time-faculty-graph}

## line graph to compare part-time faculty and non part-time faculty trends

# Recode faculty_type into two categories
staff_grouped <- staff_long %>%
  mutate(faculty_group = ifelse(faculty_type == "Part-Time Faculty", 
                                "Part-Time Faculty", 
                                "Non-Part-Time Faculty")) %>%
  group_by(year, faculty_group) %>%
  summarize(value = sum(value), .groups = "drop") 

# Plot only two groups: Part-Time vs. Non-Part-Time
ggplot(staff_grouped, aes(x = year, y = value, color = faculty_group, group = faculty_group)) +
  geom_line(size = 1) +
  ggtitle("Trends in Part-Time vs. Non-Part-Time Faculty (1975-2011)") +
  labs(x = "Year", y = "Percent of Total Instructional Staff", color = "Faculty Type") +
  theme_light()

```

### Fisheries Data

The given graphs are hard to read, because there are some huge outliers, making the rest data hard to read, especially in a 3D pie chart, where the slices are so small. I think a horizontal bar chart will improve the readability. It would also be more readable if not all countries are listed (e.g., only top 10 are listed; rest can be displayed in a table). In addition, the unit ton seems to be small for the amount of capture, using a larger unit or a log number would improve the readability. 

```{r fieshery-graph-improve}

fisheries <- read_csv("data/fisheries.csv")

## create hundred thousand ton unit variable & arrange by most capture

fisheries_capture <- fisheries %>% 
  mutate(capture_100_thousand = capture / 1e5) %>%
  arrange(desc(capture_100_thousand)) %>%
  top_n(10, capture_100_thousand) 
  

## create the horizontal bar chart

fisheries_capture %>%
  ggplot(aes(
    y = reorder(country, capture_100_thousand),  # Ensure countries are ordered by capture
    x = capture_100_thousand, 
    fill = capture_100_thousand
  )) +
  geom_col() +  
  ggtitle("Capture by Country, 2016 (in hundred thousand tons)") +
  labs(x = "Fish Capture (Hundred Thousand Tons)", y = "Country") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") + 
  theme_light()

## create hundred thousand ton unit variable & arrange by most aquaculture

fisheries_aqua <- fisheries %>% 
  mutate(aqua_100_thousand = aquaculture / 1e5) %>%
  arrange(desc(aqua_100_thousand)) %>%
  top_n(10, aqua_100_thousand) 
  

## create the horizontal bar chart

fisheries_aqua %>%
  ggplot(aes(
    y = reorder(country, aqua_100_thousand),  # Ensure countries are ordered by capture
    x = aqua_100_thousand, 
    fill = aqua_100_thousand
  )) +
  geom_col() +  
  ggtitle("Aguaculture by Country, 2016 (in hundred thousand tons)") +
  labs(x = "Fish Aquaculture (Hundred Thousand Tons)", y = "Country") +
  scale_fill_gradient(low = "lightblue", high = "darkblue") + 
  theme_light()

```
### Smokers in Whickham Data

```{r smoker-data}

data(Whickham)
??Whickham
library(performance)

```

## Excercise 1, 2, and 3

The data is most likely observational, because no control/randomly assigned conditions are reported.
There are 1314 observations in the data, each representing a participant.
There are 3 variables, representing their smoker status (categorical), age (continuous), and whether they are alive/dead (categorical).
```{r smoker-excercise-3}

Whickham %>%
  ggplot(aes(
  x = smoker,
  color = smoker,
  fill = smoker
  )) +
  geom_bar() +
  ggtitle("Smoker or Non-smoker?") +
  theme_minimal()

Whickham %>%
  ggplot(aes(
  x = outcome,
  color = outcome,
  fill = outcome
  )) +
  geom_bar() +
  ggtitle("Life Outcome") +
  theme_minimal()

Whickham %>%
  ggplot(aes(x = age)) + 
  geom_histogram(binwidth = 5, fill = "steelblue", color = "black", alpha = 0.7) + 
  ggtitle("Age Distribution of Participants") + 
  labs(x = "Age (Years)", y = "Number of Participants") + 
  theme_minimal()

```
## Excercise 4 and 5

Just based on the count, there are more alive non-smokers than smokers. I would probably expect that there is a relationship between smoker and health outcome such that being a smoker has worse health outcome than the non-smoker. However, from the graph, it is possible that the number differences between alive non-smokers vs. smokers are due to the fact that there are more alive participants in general. Therefore, I would say this relationship also needs to be examined in relation to age, where maybe smokers die earlier than non-smokers.

```{r exercise-5}

Whickham %>%
  count(smoker, outcome)

## graph the relationship between smoker status and life outcome

Whickham %>%
  ggplot(aes(
  x = outcome,
  color = smoker,
  fill = smoker
  )) + 
  geom_bar(position = "dodge") +
  labs(x = "Life outcome", y = "Count") +
  ggtitle("Smoker vs. Non-smoker Life Outcome") +
  theme_minimal()


```
## Excercise 6 and 7

In the graph, for the 18-44 group, most individuals are alive, with minimal differences between smokers and non-smokers. For the 45-64 group, death increases, and a higher proportion of smokers are deceased compared to the younger group. In the 65+ group, deaths are more common, especially among smokers, suggesting a stronger long-term impact of smoking on health. This shift can be explained by age-related mortality risks and the cumulative effects of smoking over time. The contingency table shows a consistent pattern.

```{r excercise-7}

## add the new variable

Whickham <- Whickham %>%
  mutate(age_cat = case_when(
    age <= 44 ~ "18-44",
    age > 44 & age <= 64 ~ "45-64",
    age > 64 ~ "65+"
    ))

## create a graph for all three variables

Whickham %>%
  count(smoker, age_cat, outcome)

Whickham %>%
  ggplot(aes(
  x = outcome,
  color = smoker,
  fill = smoker
  )) + 
  geom_bar(position = "dodge") +
  facet_wrap(~ age_cat) +
  labs(x = "Life outcome", y = "Count") +
  ggtitle("Smoker vs. Non-smoker Life Outcome") +
  theme_minimal()


```

